Let's Get Movin': Frontiers of Optimal Transport

### 2018-05-05 reports/paper/paper.tex
fixed first half

This project should serve as an introductory note and broad overview of both
 theoretical and computational aspects of the
 optimal transport literature. We hope that this note demonstrates both the
 theoretical elegance and practical versatility of optimal transport
 problems, and we hope to bring the reader to the forefront of the
 literature by going through a few recent articles in detail.

The organization of this report is as follows: \cref{sub:problem_intro}
introduces the discrete optimal transport problem and discusses its
interpration, dual problem, and basic theoretical properties as a linear
program. \cref{sec:exact} discusses methods to solve the optimal transport
problem \eqref{eq:ot_problem} exactly, introducing the Hungarian algorithm
\cite{kuhn2010hungarian} in \cref{sub:hungarian_algorithm} and presenting an
empirical analysis of its runtime. The Hungarian algorithm appears cubic in
run-time, confirming theoretical asymptotic analysis, which is optimal for exact
solution of \eqref{eq:ot_problem}, but is often too slow for practical
applications discussed in \cref{sub:applications}. As such, we discuss
approximate optimization methods in \cref{sec:inexact_methods}, where
\cref{sub:entropic_regularization_and_sinkhorn_s_algorithm} focuses on the
Sinkhorn-Knopp algorithm used to solve a convexification of
\eqref{eq:ot_problem} known as \ref{eq:entropic}. We present an elegant and
recent upper bound of Sinkhorn-Knopp's run-time by \cite{altschuler2017near}, as
well as numerical experiments and speculations about future work. In
\cref{sub:genetic_algorithms_}, we attempt to use genetic algorithms to
approximately solve \eqref{eq:ot_problem}, motivated by recent advances by
\cite{such2017deep}. In \cref{sec:applications_}, we showcase toy examples of
using optimal transport in computer vision with the well-known MNIST dataset.
\cref{sec:conclusion} concludes.

@@

holds, which means that $\b P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
vice versa. 
\subsection{Applications} % (fold)
\label{sub:applications}

Recent advances in machine learning, computer vision, and statistics have
 brought optimal transport problems to significant relevance. Specifically,
 Wasserstein distances have received intense attention from the computer vision
 and statistics communities. For instance, if we consider images as a discrete
 probabilistic distribution over $\mathbb{R}^2$, the Wasserstein distance turns
 out to mimick human perceptions of similarity: Images that are similar to
 humans are close in Wasserstein distances (See \cite{solomon2015convolutional}
 for an illustration of the power of Wasserstein metrics). Other distances over
 distributions, such as the well-known Kullback-Leibler divergence, do not have
 such properties. Such properties of the Wasserstein distance also motivates
 advances in image generation via an incorporation with the increasingly popular
 generative adversarial network (GAN) framework \cite{arjovsky2017wasserstein}.

@@

The \ref{eq:ot_problem} Optimal Transport problem as formulated is an LP. This
means that a natural first-step to benchmarking the problem difficulty is to use
a generic LP solver. The only difficulty we face when doing this is that we must
flatten our arguments from matrices to vectors in order to fit the API of a
generic LP solver, such as \texttt{cvxopt.solvers.lp} which we employ. Recall
that our problem inputs are $\mathbf{a} \in \mathbb{R}^n_{\ge 0}$, $\mathbf{b}
\in \mathbb{R}^m_{\ge 0}$ with sum of entries equal to 1 and $\mathbf{C} \in
\mathbb{R}^{n,m}$. We flatten $C$ into a vector $\mathbf{c} \in \mathbb{R}^{nm}$
and define the matrix
\[ \mathbf{A} := \begin{bmatrix} I_n \otimes \mathbf{1}_m \\ \mathbf{1}_n
\otimes I_m \end{bmatrix} \]
which allows us to formulate our LP as,
\[ \min \mathbf{c}^\top \mathbf{x} \text{ such that } \mathbf{x} \ge 0,
\mathbf{A} \mathbf{x} = \begin{bmatrix} \mathbf{a} \\ \mathbf{b}
\end{bmatrix}.\]

@@

One such special case of the optimal transport problem is when $\mathbf{a} =
\mathbf{b} \propto \mathbf{1}_n$ as in our simulations to test the performance
of the generic LP solver. In this case, mass is being matched in unit pieces,
which makes the problem more tractable. The Hungarian algorithm solves this
special-case. The Hungarian algorithm is also called a dual-ascent method
because it starts by establishing feasible solutions to the primal and dual
problems, and proceeds by improving these solutions until their objectives
match. We empirically test the performance of such an algorithm, providing the
results in \cref{fig:hungarianruntime}. As we can note when comparing to the
general case LP, the run-times for comparably sized problems are distinctly
smaller for the Hungarian algorithm.

>>>>

This project should serve as an introductory note and broad overview of both
 the theoretical and the computational aspects of the
 optimal transport literature. We hope that this note demonstrates the
 theoretical elegance and practical versatility of optimal transport
 problems, and we hope to bring the reader to the forefront of the
 literature by going through a few recent articles in detail.

The organization of this report is as follows: \cref{sub:problem_intro}
introduces the discrete optimal transport problem and discusses its
interpration, dual problem, and basic theoretical properties as a linear
program. \cref{sec:exact} discusses methods to solve the optimal transport
problem \eqref{eq:ot_problem} exactly, introducing the Hungarian algorithm
\cite{kuhn2010hungarian} in \cref{sub:hungarian_algorithm} and presenting an
empirical analysis of its runtime. The Hungarian algorithm appears cubic in
run-time, confirming theoretical asymptotic analysis; the cubic run-time 
is also optimal for exact
solution of \eqref{eq:ot_problem}, but is often too slow for practical
applications discussed in \cref{sub:applications}. As such, we discuss
approximate optimization methods in \cref{sec:inexact_methods}, where
\cref{sub:entropic_regularization_and_sinkhorn_s_algorithm} focuses on the
Sinkhorn-Knopp algorithm used to solve a convexification of
\eqref{eq:ot_problem} known as \ref{eq:entropic}. We present an elegant and
recent upper bound of Sinkhorn-Knopp's run-time by \cite{altschuler2017near}, as
well as numerical experiments and speculations about future work. In
\cref{sub:genetic_algorithms_}, we attempt to use genetic algorithms to
approximately solve \eqref{eq:ot_problem}, motivated by recent advances by
\cite{such2017deep}. In \cref{sec:applications_}, we showcase toy examples of
using optimal transport in computer vision with the well-known MNIST dataset.
\cref{sec:conclusion} concludes.

@@

holds, which means that $P_{ij} \neq 0 \implies C_{ij} = f_i - g_j$ and
$C_{ij} \neq f_i - g_j \implies P_{ij} = 0$. 
\subsection{Applications} % (fold)
\label{sub:applications}

Recent advances in machine learning, computer vision, and statistics have
 brought optimal transport problems to significant relevance. Specifically,
 Wasserstein distances have received intense attention from the computer vision
 and statistics communities. For instance, if we consider an image as a discrete
 probabilisty distribution over $\mathbb{R}^2$, where the intensity of a pixel represents density, the Wasserstein distance turns
 out to mimick human perceptions of similarity: Images that are similar to
 humans are close in Wasserstein distances (See \cite{solomon2015convolutional}
 for an illustration of the power of Wasserstein metrics). Other distances over
 distributions, such as the well-known Kullback-Leibler divergence, do not have
 such properties. Such properties of the Wasserstein distance also motivates
 advances in image generation via an incorporation with the increasingly popular
 generative adversarial network (GAN) framework \cite{arjovsky2017wasserstein}.

@@

The \ref{eq:ot_problem} Optimal Transport problem, as formulated, is an LP. This
means that a natural first-step to benchmarking the problem difficulty is to use
a generic LP solver. The only difficulty we face when doing this is that we must
flatten our arguments from matrices to vectors in order to fit the API of a
generic LP solver, such as \texttt{cvxopt.solvers.lp} which we employ. Recall
that our problem inputs are $\mathbf{a} \in \mathbb{R}^n_{\ge 0}$, $\mathbf{b}
\in \mathbb{R}^m_{\ge 0}$ with sum of entries equal to 1 and $\mathbf{C} \in
\mathbb{R}^{n,m}$. We flatten $C$ into a vector $\mathbf{c} \in \mathbb{R}^{nm}$
and define the matrix
\[ \mathbf{A} := \begin{bmatrix} I_n \otimes \mathbf{1}_m \\ \mathbf{1}_n
\otimes I_m \end{bmatrix} \]
which allows us to formulate our LP as,
\[ \min \mathbf{c}^\top \mathbf{x} \text{ such that } \mathbf{x} \ge 0,
\mathbf{A} \mathbf{x} = \begin{bmatrix} \mathbf{a} \\ \mathbf{b}
\end{bmatrix}.\]

@@

One such special case of the optimal transport problem is when $\mathbf{a} =
\mathbf{b} \propto \mathbf{1}_n$ as in our simulations to test the performance
of the generic LP solver. In this case, mass is being matched in unit pieces,
which makes the problem more tractable. The Hungarian algorithm solves this
special-case. The Hungarian algorithm is also called a dual-ascent method
because it starts by establishing feasible solutions to the primal and dual
problems, and proceeds by improving these solutions until their objectives
match. We empirically test the performance of such an algorithm, providing the
results in \cref{fig:hungarianruntime}. As we can note when comparing to the
general case LP, the run-times for comparably sized problems are distinctly
smaller for the Hungarian algorithm. For details about the Hungarian algorithm, see \citep[Chapter 3.6]{peyre2017computational}.
